---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# RBATS

> The `RBATS` is a experimental R package for Bayesian Dynamic Models. The functionaly is based on the
[Bayesian Forecasting and Dynamic Models](http://www2.stat.duke.edu/~mw/West&HarrisonBook/)
from West and Harrison (1997). The currently version of the package implements the Dynamic Linear Models and Dynamic Generalized Exponential Growth Models.


## Development cycle

This package is early in development. Changes and additional functionality may occur.
The **development** version can be installed from GitHub using:
``` r
# install.packages("remotes")
remotes::install_github("AndrMenezes/RBATS", build_vignettes = TRUE)
```

## Quick overview

The package is structured based on S3 classes objects. 
The following functions created the two main objects of the package:

- `dlm`: create and object of class `dlm` for Dynamic Linear Models. 

- `dgegm`: create and object of class `dgegm` for Dynamic Generalized Exponential Growth Models.

Each of this class objects have the following methods:

- `update_moments`: update the moments of state parameters for `dlm` and `dgegm` objects.

- `retrospective_moments`: smoothing moments of state parameters for `dlm` objects.

- `fit`: perform the filtering and smoothing for `dlm` and `dgegm` objects.

- `forecast`: perform the marginal forecast for `dlm.fit` objects.

- `logLik`: return the predictive log-likelihood for `dlm.fit` and `dgegm.fit` objects.

Two vignettes showing the package functionality can be accessed through:

```{r}
utils::browseVignettes(package = "RBATS")
```



## Contribute

The `RBATS` is still an experimental `R` package. I am inspired in the great
[`PyBATS`](https://lavinei.github.io/pybats/) python package from  [Isaac Lavine](https://www.linkedin.com/in/isaac-lavine-70495929/) and
[Andrew Cron](https://www.linkedin.com/in/andrewjcron/).
I hope in future `RBATS` became a robust tool as well as `PyBATS`. 

Please, any suggestions and contributions are more than welcome.
Feel free to open issues, pull requests and/or forks. Thanks!

## Other R packages

Other R packages that can be used for Dynamic Linear Model or also known
State Space Models are:

- [`dlm`](https://cran.r-project.org/web/packages/dlm/index.html) by Giovanni Petris and Wally Gilks.

- [`bsts`](https://cran.r-project.org/web/packages/bsts/index.html) by Steven L. Scott.

- [`KFAS`](https://cran.r-project.org/web/packages/KFAS/index.html) by Jouni Helske.

- [`bssm`](https://cran.r-project.org/web/packages/bssm/index.html) by Jouni Helske and Matti Vihola.




