---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.height = 6,
  fig.width = 8,
  eval = TRUE
)
```

# RBATS

> The `RBATS` is an experimental R package for Bayesian Dynamic Models. The functionaly is based on the
[Bayesian Forecasting and Dynamic Models](http://www2.stat.duke.edu/~mw/West&HarrisonBook/)
from West and Harrison (1997). The currently version of the package implements the Dynamic Linear Models and Dynamic Generalized Exponential Growth Models.


## Development cycle

This package is early in development. Changes and additional functionality may occur.
The **development** version can be installed from GitHub using:
```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("AndrMenezes/RBATS", build_vignettes = TRUE)
```

## Quick overview

The package is structured based on S3 classes objects. 
The following functions created the two main objects of the package:

- `dlm`: create and object of class `dlm` for Dynamic Linear Models. 

- `dgegm`: create and object of class `dgegm` for Dynamic Generalized Exponential Growth Models.

Each of this class objects have the following methods:

- `update_moments`: update the moments of state parameters for `dlm` and `dgegm` objects.

- `fit`: performs the filtering and smoothing for `dlm` and `dgegm` objects.

- `forward_filter`: performs the filtering for `dlm` and `dgegm` objects.

- `backward_smoother`: performs the smoothing for `dlm` and `dgegm` objects.

- `forecast`: performs the marginal forecast for `dlm` and `dgegm` objects.

- `logLik`: returns the predictive log-likelihood for `dlm` and `dgegm` objects.

- `plot`: returns base plots of filter and/or smooth predictive and posterior distribution for `dlm` and `dgegm` objects.  


First, we define the model using the `dlm` function, then we fit the model through the
`fit` method.

```{r model-fit}
library(RBATS)
(model_object <- dlm(polynomial_order = 2, discount_factors = list(polynomial = 0.95)))
(fitted_model <- fit(model_object, y = c(Nile)))
```

Basics visual inspects can be accessed using the `plot` method.
```{r plot-method}
plot(fitted_model, what = "posterior", type = "filter")
plot(fitted_model, what = "posterior", type = "smooth")
```

The `fit` method returns the `dlm` model (`model`) with the most updated prior and
posterior. Then, we can use the most recent model to obtain forecast.
```{r forecast-method}
model_object <- fitted_model$model
forecast(model_object, horizon = 5)
```

One of the main advantages of DLM are the sequential learning, which means that we can
update the model when new observation came without need the past information.
The `update_moments` method is used for this purpose. 

```{r update-method}
set.seed(1212)
y_new <- rnorm(1, mean(Nile), sd = 1)
model_object
model_object_updated <- update_moments(model_object, y = y_new)
model_object_updated
```



## Contribute

The `RBATS` is still an experimental `R` package. I am inspired in the great
[`PyBATS`](https://lavinei.github.io/pybats/) python package from  [Isaac Lavine](https://www.linkedin.com/in/isaac-lavine-70495929/) and
[Andrew Cron](https://www.linkedin.com/in/andrewjcron/).
I hope in future `RBATS` became a robust tool as well as `PyBATS`. 

Please, any suggestions and contributions are more than welcome.
Feel free to open issues, pull requests and/or forks. Thanks!

## Other R packages

Other R packages that can be used for Dynamic Linear Model or also known
State Space Models are:

- [`dlm`](https://cran.r-project.org/web/packages/dlm/index.html) by Giovanni Petris and Wally Gilks.

- [`bsts`](https://cran.r-project.org/web/packages/bsts/index.html) by Steven L. Scott.

- [`KFAS`](https://cran.r-project.org/web/packages/KFAS/index.html) by Jouni Helske.

- [`bssm`](https://cran.r-project.org/web/packages/bssm/index.html) by Jouni Helske and Matti Vihola.




